---
title: "STAT536HW4"
author: "Bryan Ng, 2427348"
date: "2025-11-08"
output: pdf_document
---

```{r, warning=FALSE}
data <- matrix(c(28, 656,
                 18, 658),
               nrow = 2, byrow = TRUE,
               dimnames = list(Treatment = c("Placebo","Aspirin"),
                               MI        = c("Yes","No")))
data
```

\section{Q1}

```{r}
row_totals <- rowSums(data)
col_totals <- colSums(data)
grand_total <- sum(data)

row_totals; col_totals; grand_total
```

\section{Q2}

```{r}
r <- rowSums(data)
c <- colSums(data)
N <- sum(data)

bounds_2x2 <- function(r, c, N){
  L11 <- max(0, r[1] + c[1] - N); U11 <- min(r[1], c[1])
  L12 <- max(0, r[1] + c[2] - N); U12 <- min(r[1], c[2])
  L21 <- max(0, r[2] + c[1] - N); U21 <- min(r[2], c[1])
  L22 <- max(0, r[2] + c[2] - N); U22 <- min(r[2], c[2])
  data.frame(cell = c("a11(Placebo-Yes)","a12(Placebo-No)",
                      "a21(Aspirin-Yes)","a22(Aspirin-No)"),
             lower = c(L11,L12,L21,L22),
             upper = c(U11,U12,U21,U22))
}
bounds_2x2(r, c, N)

```

This means the first cell, which reflects to Placebo–Yes, can vary from 0 to 46 while keeping all margins constant, and the other cells adjust accordingly.
Therefore, there are 47 valid tables that satisfy the given row and column totals.

\section{Q3}

```{r}
L11 <- max(0, r[1] + c[1] - N); U11 <- min(r[1], c[1])

num_tables <- U11 - L11 + 1
num_tables

all_tables <- data.frame(
  a11 = L11:U11,
  a12 = r[1] - (L11:U11),
  a21 = c[1] - (L11:U11),
  a22 = N - (L11:U11) - (r[1] - (L11:U11)) - (c[1] - (L11:U11))
)
head(all_tables)

```


\section{Q4}

```{r}
expected_saturated <- data
expected_saturated
```

Under the saturated model, it is clear that the expected counts match the observed counts exactly.

\section{Q5}

```{r}
chi_ind <- chisq.test(data, correct = FALSE)
expected_independence <- chi_ind$expected
expected_independence

```

Under the independence model, expected counts differ slightly from the observed values, reflecting a weak association.

\section{Q6}

```{r}
test_res <- chisq.test(data, correct = FALSE)
test_res$statistic   
test_res$parameter   
test_res$p.value    
test_res             
```

The Pearson chi-square test indicates no significant relationship between aspirin use and myocardial infarction.


\section{Q7}

```{r}
expected <- chisq.test(data, correct = FALSE)$expected

G2 <- 2 * sum(data * log(data / expected))
df <- (nrow(data) - 1) * (ncol(data) - 1)
p_value_G2 <- 1 - pchisq(G2, df)
p_value_G2
```

The likelihood-ratio test also leads to the same as the Pearson chi-square test, indicating no significant relationship between aspirin use and myocardial infarction.

\section{Q8}

```{r}
fisher_test <- fisher.test(data, alternative = "two.sided")
fisher_test
```

Fisher’s exact test also fails to reject independence; the odds ratio confidence interval includes 1.

\section{Q9}

```{r}
odds_yes <- data["Placebo","Yes"] / data["Aspirin","Yes"]
odds_no  <- data["Placebo","No"]  / data["Aspirin","No"]

log_ratio_yes <- log(odds_yes)
log_ratio_no  <- log(odds_no)

cat("log odds ratio given MI=Yes:", log_ratio_yes, "\n")
cat("log odds ratio given MI=No :", log_ratio_no, "\n")
```

The log-odds ratio shows that the placebo group has slightly higher MI odds.

\section{Q10}

Overall, all tests suggest that aspirin use is not significantly associated with myocardial infarction in this dataset.
